Ab Initio Effective One-Electron Potential Operators: Applications for Charge-Transfer Energy in Effective Fragment Potentials

### 2019-11-16 papers/2_oep_technique/paper-r2.tex
New fogures added and EFP2 equation initiated

Charge transfer (CT) between molecules occurs when the net electronic populations 
of interacting molecules change which leads to an additional stabilization 
of a molecular aggregate. The effect of CT on the interaction energy 
is known to be often non\hyp{}negligible, especially
in donor\hyp{}acceptor systems such as H\hyp{}bonded species
and charged complexes.
\emph{Ab initio} fragmentation methods, such as
the second generation of the effective fragment potential method (EFP2),
are often used to describe the homogeneous extended molecular systems 
such as solutions
and recently even biomolecules. EFP2 relies on the intermolecular 
perturbation theory, and the total interaction energy is approximated as
a sum of five contributions due to the Coulombic interaction of the unperturbed (isolated)
charge densities of the monomers, exchange\hyp{}repulsion energy,
induction energy due to mutual polarization, intermolecular dispersion energy,
as well as the CT energy,
%
\begin{equation}
 U^{\rm EFP2} \approx U^{\rm Coul} + U^{\rm Ex-Rep} + U^{\rm Ind} + U^{\rm Disp} + U^{\rm CT} \;.
\end{equation}
%
However, calculating the CT stabilization energy is usually much more expensive 
than evaluating the contributions due to other effects.\cite{Gordon.Fedorov.Pruitt.Slipchenko.ChemRev.2012}
%However, the computational cost of the CT term is usually at least few times larger
%than the next costly contribution $U^{\rm Ex-Rep}$, which is still too costly
%to perform molecular dynamics (MD) simulations including the $U^{\rm CT}$. 
In effect,
this part of the EFP2 energy is usually ignored in most of the applications. In the 
recent LIBEFP library, CT term were even not implemented due to those reasons.

@@

\paragraph{Group (ii).}
The term belonging to this group can be considered as a sum of interaction
energies between the total charge density distribution of molecule $B$
and the partial density $\rho_{ik}({\bf r})$ of molecule $A$,
weighted by the overlap integrals $S_{nk}$. Using the distributed multipole 
expansion based on the charge centroids of the localized molecular orbitals (LMO's),
${\bf r}_i = \tBraKet{i}{\hat{\bf r}}{i}$ with $\chi_i({\bf r})$ being localized,
this group can be written as
%
\begin{equation} \label{e:ct.group-2}
      - \sum_{k\in A}^{\rm Occ} S_{nk}  \tBraKet{k}{\hat{v}^B_{\rm tot} }{i} 
 \cong - \sum_{k\in A}^{\rm Occ} S_{nk} \rho_{ki}^A \odot \rho^B_{\rm tot} \;.
\end{equation}
%
Note that, in general, $\rho_{ki}^A$ is complex whereas $\rho^B_{\rm tot}$ is always real.
Since the interaction energy must also be real, it must hold that
%
\begin{equation} \label{e:ct.group-2-d}
 \rho_{ki}^A \odot \rho^B_{\rm tot} = \delta_{ik} \rho_{ii}^A \odot \rho^B_{\rm tot} \;.
\end{equation}
%
Consider now a situation in which the localized occupied MO's are used.
In this case, the distributed multipole expansion
of the above becomes
%
\begin{multline}
 \rho_{ii}^A \odot \rho^B_{\rm tot} 
 \approx 
 q_{i} 
 \left[
  %\sum_{i\in A}^{\rm LC} 
 \sum_{y\in B}^{\rm At}
  \frac{Z_y}{\vert {\bf r}_y - {\bf r}_i \vert } 
 +
 2\sum_{j\in B}^{\rm LC}
  \frac{q_j}{\vert {\bf r}_j - {\bf r}_i \vert } 
 \right] \\
 + \text{ quadrupole terms } \;,
\end{multline}
%
because the distributed charges
$q_i = -1$ whereas the distributed dipole moments %${\BM{\upmu}_i}$ 
centered at their
respective LMO charge centroids vanish.
This means that Eq.~\eqref{e:ct.group-2} can be approximated as follows:
%
\begin{equation} \label{e:ct.group-ii.final}
      - \sum_{k\in A}^{\rm Occ} S_{nk} \tBraKet{k}{\hat{v}^B_{\rm tot} }{i} 
 \approx S_{ni}  \left[
  %\sum_{i\in A}^{\rm LC} 
 \sum_{y\in B}^{\rm At}
  \frac{Z_y}{ r_{yi} } 
 -
 \sum_{j\in B}^{\rm LC}
  \frac{2}{r_{ji}} 
 \right] \;.
\end{equation}
%
Therefore, only overlap integrals and relative distances between
atomic and LC's are needed, which leads to a great reduction of the
calculation cost,
as compared either to the original expression or to the multipole expansion (left\hyp{} and right\hyp{}hand sides
of Eq.~\eqref{e:ct.group-2}, respectively).
We shall refer to this approximation as to
the spherical overlap approximation (SOA). Note that in order for this approximation to be valid,
occupied molecular orbitals need to be localized.

@@

\subsection{\label{s:gsdzdv}Accuracy of OEP-based model}
%
\begin{figure}[t]
\includegraphics[width=0.5\textwidth]{fig-2.eps}
\caption{\label{f:fig-1} {\bf Accuracy of the OEP approach to estimate the charge transfer energy
in the PT/HF formulation.} 
Asymptotic dependencies of the charge transfer energy is plotted for 
(a) water dimer, 
(b) water\hyp{}ammonium complex, and 
(c) ammonium\hyp{}nitrate complex,
where one molecule has been translated by $R$ from the starting geometry
along the vector specified in Fig. 1.
The counterpoise\hyp{}corrected total interaction energy
is also shown for comparison in light grey color in this figure.
All data were obtained at HF/6-31+G** level of theory.
} 
\end{figure}
%

>>>>

Charge transfer (CT) between molecules occurs when the net electronic populations 
of interacting molecules change which leads to an additional stabilization 
of a molecular aggregate. The effect of CT on the interaction energy 
is known to be often non\hyp{}negligible, especially
in donor\hyp{}acceptor systems such as H\hyp{}bonded species
and charged complexes.
\emph{Ab initio} fragmentation methods, such as
the second generation of the effective fragment potential method (EFP2),
are often used to describe the homogeneous extended molecular systems 
such as solutions
and recently even biomolecules. EFP2 relies on the intermolecular 
perturbation theory, and the total interaction energy is approximated as
a sum of five contributions due to the Coulombic interaction of the unperturbed (isolated)
charge densities of the monomers, exchange\hyp{}repulsion energy,
induction energy due to mutual polarization, intermolecular dispersion energy,
as well as the CT energy,
%
\begin{equation}
 U^{\rm EFP2} \approx U^{\rm Coul} + U^{\rm Ex-Rep} + U^{\rm Ind} + U^{\rm Disp} + U^{\rm CT} \;.
\end{equation}
%
Xu and Gordon used the expansion of the overlap density 
in Taylor series and found four different approximate
theories for the CT energy. The optimal theory, which was shown
by them to best reproduce the reduced variational scheme (RVS)
reads
%
\begin{equation}
 E^{A\rightarrow B} \approx 
 2\sum_{i\in A}^{\rm Occ}
  \sum_{n\in B}^{\rm Vir}
   \frac{ \left| V_{in}^{A\rightarrow B} \right|^2}{F_{ii} - T_{nn}} \;,
\end{equation}
%
where
%
\begin{multline}
 \left| V_{in}^{A\rightarrow B} \right|^2 \approx
 \frac{U_{in}}{1-\sum_{m\in A}^{\rm All} S_{mn}^2}
 \Bigg\{
   U_{in} \\ + \sum_{j\in B}^{\rm Occ} S_{ij} 
   \left( 
     T_{nj} - \sum_{m\in A}^{\rm All} S_{nm} T_{mj}
   \right)
 \Bigg\} \;,
\end{multline}
%
and 
%
\begin{equation}
 U_{in} \equiv V_{in}^{{\rm EF},B} - \sum_{m\in A}^{\rm All} V_{im}^{{\rm EF},B} S_{mn} \;.
\end{equation}
%
However, calculating the CT stabilization energy is usually much more expensive 
than evaluating the contributions due to other effects.\cite{Gordon.Fedorov.Pruitt.Slipchenko.ChemRev.2012}
%However, the computational cost of the CT term is usually at least few times larger
%than the next costly contribution $U^{\rm Ex-Rep}$, which is still too costly
%to perform molecular dynamics (MD) simulations including the $U^{\rm CT}$. 
In effect,
this part of the EFP2 energy is usually ignored in most of the applications. In the 
recent LIBEFP library, CT term were even not implemented due to those reasons.

@@

\paragraph{Group (ii).}
The term belonging to this group can be considered as a sum of interaction
energies between the total charge density distribution of molecule $B$
and the partial density $\rho_{ik}({\bf r})$ of molecule $A$,
weighted by the overlap integrals $S_{nk}$. Using the distributed multipole 
expansion based on the charge centroids of the localized molecular orbitals (LMO's),
${\bf r}_i = \tBraKet{i}{\hat{\bf r}}{i}$ with $\chi_i({\bf r})$ being localized,
this group can be written as
%
\begin{equation} \label{e:ct.group-2}
      - \sum_{k\in A}^{\rm Occ} S_{nk}  \tBraKet{k}{\hat{v}^B_{\rm tot} }{i} 
 \cong - \sum_{k\in A}^{\rm Occ} S_{nk} \rho_{ki}^A \odot \rho^B_{\rm tot} \;.
\end{equation}
%
Note that, in general, $\rho_{ki}^A$ is complex whereas $\rho^B_{\rm tot}$ is always real.
Since the interaction energy must also be real, it must hold that
%
\begin{equation} \label{e:ct.group-2-d}
 \rho_{ki}^A \odot \rho^B_{\rm tot} = \delta_{ik} \rho_{ii}^A \odot \rho^B_{\rm tot} \;.
\end{equation}
%
Consider now a situation in which the localized occupied MO's are used.
In this case, the distributed multipole expansion
of the above becomes
%
\begin{multline}
 \rho_{ii}^A \odot \rho^B_{\rm tot} 
 \approx 
 q_{i} 
 \left[
  %\sum_{i\in A}^{\rm LC} 
 \sum_{y\in B}^{\rm At}
  \frac{Z_y}{\vert {\bf r}_y - {\bf r}_i \vert } 
 +
 2\sum_{j\in B}^{\rm LC}
  \frac{q_j}{\vert {\bf r}_j - {\bf r}_i \vert } 
 \right] \\
 + \text{ quadrupole terms } \;,
\end{multline}
%
because the distributed charges
$q_i = -1$ whereas the distributed dipole moments %${\BM{\upmu}_i}$ 
centered at their
respective LMO charge centroids vanish.\cite{Etchebest.Lavery.Pullman.TheorChimActa.1982}
This means that Eq.~\eqref{e:ct.group-2} can be approximated as follows:
%
\begin{equation} \label{e:ct.group-ii.final}
      - \sum_{k\in A}^{\rm Occ} S_{nk} \tBraKet{k}{\hat{v}^B_{\rm tot} }{i} 
 \approx S_{ni}  \left[
  %\sum_{i\in A}^{\rm LC} 
 \sum_{y\in B}^{\rm At}
  \frac{Z_y}{ r_{yi} } 
 -
 \sum_{j\in B}^{\rm LC}
  \frac{2}{r_{ji}} 
 \right] \;.
\end{equation}
%
Therefore, only overlap integrals and relative distances between
atomic and LC's are needed, which leads to a great reduction of the
calculation cost,
as compared either to the original expression or to the multipole expansion (left\hyp{} and right\hyp{}hand sides
of Eq.~\eqref{e:ct.group-2}, respectively).
We shall refer to this approximation as to
the spherical overlap approximation (SOA). Note that in order for this approximation to be valid,
occupied molecular orbitals need to be localized.

@@

\subsection{\label{s:gsdzdv}Accuracy of OEP-based model}

First, the asymptotic dependence of the CT energy
was investigated for four chosen molecular pairs and the results are
shown in Figure~\ref{f:fig-1} and Figure~S1 for the methanol-water system. 
Overall, OEP technique of effective
elimination of ERI's sketched in Eq.~\eqref{e:ft-reduction} 
works
quantitatively well for water dimer and ammonium nitrate (Figures~\ref{f:fig-1}(a) and (c)),
and methanol-water (Figure~S1)
reproducing the reference Otto\hyp{}Ladik CT energy,
whereas in the case of water\hyp{}ammonium cation complex the OEP
method overestimates the CT energy by around 50\%, even for large
intermonomer separations. Careful inspection
(see SI)
revealed that the reason is the SOA in $G_2$ and $G_3$ and not
the generalized density fitting in $G_1$ terms, in the case of which
the aug-cc-pVDZ-jkfit basis set was sufficient as auxiliary space of OEP transformation.
On the other hand, analysis of the statistical sets of dimeric complexes
shown in Figure~XXX
confirmed that in the great majority of cases the 
OEP technique gives reasonably accurate results with respect
to the parent model with relative errors less than 10\%,
unless systems are charged.

In the case of water dimer and methanol-water complex, EFP2 and PT/HF estimations
coincide with the OEP model. However, in the presence of ions, 
the deviation of the OEP model from the PT/HF energy 
become significant, in contrast to the EFP2 model, which gives very good estimates
of the PT/HF energy in mostly all cases.
Naturally, the accuracy of the Otto\hyp{}Ladik model limits the absolute accuracy
of the OEP model, which is manifested in the ammonium nitrate system
in Figure~\ref{f:fig-1}(c) or the problems with SOA. 
In order to improve the performance for ionic systems,
further studies are necessary, that will modify the original formulation
of Otto and Ladik and SOA treatment. Nevertheless, except from the ionic systems,
it is clear that the level of accuracy of the EFP2 and OEP models are rather comparable
(Figure~XX) across a wide range of various interacting complexes including 
systems with H-boding, dominated by dipole-dipole or dispersion interactions or 
exhibiting substantial CT.

%
\begin{figure}[h]
\includegraphics[width=0.46\textwidth]{fig-2.eps}
\caption{\label{f:fig-1} {\bf Asymptotic dependence of the charge transfer energy
in the PT/HF formulation for selected bi\hyp{}molecular complexes.} 
The charge transfer energy estimates are plotted for 
(a) water dimer, 
(b) water\hyp{}ammonium complex, and 
(c) ammonium\hyp{}nitrate complex,
where one molecule has been translated by $R$ from the starting geometry
along the vector specified in Fig. 1.
The counterpoise\hyp{}corrected total interaction energy
is also shown for comparison in light grey color in this figure.
All data were obtained at HF/6-31+G** level of theory.
} 
\end{figure}
%

%
\begin{figure*}[h]
\includegraphics[width=0.8\textwidth]{fig-3.eps}
\caption{\label{f:fig-1} {\bf Performance of the OEP-based vs EFP2 method for the PT/HF charge transfer interaction energy.}
(a) accuracy of the ERI elimination technique; 
(b) accuracy of the OEP and EFP2 estimates of the CT energy with respect to the
PT/HF estimate. 
The structures of bi\hyp{}molecular complexes
were taken from the databases developed by the Truhlar's group:
the HB6/04 hydrogen bonding database 
(red),\cite{Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
the DI6/04 dipoleinteraction database 
(green),\cite{Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
the CT7/04 charge-transfer complex database 
(blue),\cite{Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
the WI7/05 weak interaction database 
(yellow),\cite{Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005,Zhao.Truhlar.JPCA.2005}
and PPS5/05 the $\pi$-$\pi$ stacking database 
(black).\cite{Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005,Zhao.Truhlar.JPCA.2005}
6-311++G** basis set was used for the CT energy calculations.
} 
\end{figure*}
%
